---
title: 'Data Analytics and ML: Midterm Exam'
author: "Pongsun B."
date: "2026-03-11"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if(!require(tidyverse)){install.packages("tidyverse")}
if(!require(ggbeeswarm)){install.packages("ggbeeswarm", dependencies = TRUE)}
if(!require(moments)){install.packages("moments")}
if(!require(corrplot)){install.packages("corrplot")}
if(!require(mlr)){install.packages("mlr", dependencies = TRUE)}
if(!require(caret)){install.packages("caret")}
if(!require(mlbench)){install.packages("mlbench")}
if(!require(rpart.plot)){install.packages("rpart.plot")}
if(!require(kernlab)){install.packages("kernlab")}
if(!require(FSelectorRcpp)){install.packages("FSelectorRcpp")}

library(parallelMap)
library(parallel)
```

# Optimize the Loan Eligibility Process

Dream Housing Finance company deals in all kinds of home loans...

## Data Dictionary

...

```{r Getting and Exploring the Data, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}

myURL = "https://www.dropbox.com/scl/fi/wn5ow4zvgg4p4odfbor51/LoanPrediction.csv?rlkey=mludfatyg87sgz8374i0h8kfa&st=6kivxmtg&dl=1"

myData = as_tibble(read_csv(myURL, na = c("", " ", NA), show_col_types = FALSE))
myData

mlr::summarizeColumns(myData)
```

## Data Manipulation

```{r Data Manipulation, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}

# Step 1
myData <- myData |>
  mutate(across(where(is.character) & !matches("Loan_ID"), as.factor))

myData <- myData |>
  mutate(Credit_History = factor(Credit_History, levels = c(0, 1), labels = c("No", "Yes")))

# Step 2
need_check = c("LoanAmount", "ApplicantIncome", "CoapplicantIncome")

skewness_values <- sapply(myData[, need_check], moments::skewness, na.rm = TRUE)
print(skewness_values)

myData |>
  select(all_of(need_check)) |>
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") |>
  ggplot(aes(x = Value, fill = Variable)) +
  geom_histogram(bins = 30, color = "white") +
  facet_grid(. ~ Variable, scales = "free_x") +
  labs(x = "Value", y = "count") +
  theme_bw()

# Shapiro-Wilk normality test
lapply(myData[, need_check], function(x) shapiro.test(na.omit(x)))
```

## Missing Value Imputation

```{r Missing Value Imputation, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}

# Impute missing values by mean and mode
myData_imp <- mlr::impute(myData, 
                     classes = list(factor = imputeMode(), 
                                    numeric = imputeMean()))

myData_imputed <- myData_imp$data

myData_imputed$Loan_ID <- NULL

summarizeColumns(myData_imputed[,1:12])
```

## Feature Engineering

```{r Feature Engineering, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}

myData_trim <- myData_imputed |> 
  capLargeValues(target = "Loan_Status", cols = c("ApplicantIncome"), threshold = 40000) |> 
  capLargeValues(target = "Loan_Status", cols = c("CoapplicantIncome"), threshold = 21000) |> 
  capLargeValues(target = "Loan_Status", cols = c("LoanAmount"), threshold = 520)

# Step 3: Create new features
myData_added <- myData_trim |>
  mutate(
    Total_Income        = ApplicantIncome + CoapplicantIncome,
    Loan_by_Income      = LoanAmount / Total_Income,
    Loan_Amount_by_Term = LoanAmount / Loan_Amount_Term
  )

myData_corr <- myData_added |> select(where(is.numeric))

cor_matrix <- cor(myData_corr, use = "complete.obs")

corrplot(cor_matrix,
         method  = "number",
         type    = "upper",
         diag    = FALSE,
         number.cex = 0.7,
         tl.cex  = 0.8,
         addCoef.col = "black",
         number.digits = 2)

myData_final <- myData_added |> 
  mutate(Total_Income = NULL,
         Loan_ID = as_factor(myData$Loan_ID))
```

## Data Visualization

```{r Data Visualization, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}

# Step 4

long_data_cat <- myData_final |>
  select(where(is.factor)) |>
  mutate_at(.vars = vars(-Loan_ID, -Loan_Status),
            .funs = as.integer) |>
  pivot_longer(names_to  = "Variable",
               values_to = "Value",
               -c(Loan_ID, Loan_Status))

long_data_cat |>
  ggplot(aes(x = Value, fill = Loan_Status)) +
  geom_bar(position = "fill") +
  facet_wrap(~ Variable, scales = "free_x") +
  labs(y = "count") +
  theme_bw()

long_data_num <- myData_final |>
  select(Loan_ID, Loan_Status, where(is.numeric)) |>
  pivot_longer(names_to  = "Variable",
               values_to = "Value",
               -c(Loan_ID, Loan_Status))

long_data_num |>
  ggplot(aes(x = Loan_Status, y = Value, fill = Loan_Status)) +
  geom_boxplot(outlier.colour = "red", outlier.shape = 16) +
  facet_wrap(~ Variable, scales = "free_y") +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "Loan_Status", y = "Value")
```

## Hypothesis Testing

```{r Hypothesis Testing, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}

# Step 5

chi_results <- long_data_cat |>
  split(f = long_data_cat$Variable) |>
  lapply(function(df) chisq.test(table(df$Value, df$Loan_Status)))

cat("Chi-Square p-values:\n")
sapply(chi_results, function(x) x$p.value)

kw_results <- long_data_num |>
  split(f = long_data_num$Variable) |>
  lapply(function(df) kruskal.test(Value ~ Loan_Status, data = df))

cat("\nKruskal-Wallis p-values:\n")
sapply(kw_results, function(x) x$p.value)
```

## Create the Task

```{r Create a Task, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}

myData_train <- myData_final |>
  select(-Loan_ID)

myTask <- makeClassifTask(data = myData_train, target = "Loan_Status")
myTask <- normalizeFeatures(myTask)
str(getTaskData(myTask))
```

## Discriminant Analysis

```{r Discriminant Analysis, message=FALSE, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}

# ── LDA ─────────────────────────────────────────────────────────────────────
lda <- makeLearner("classif.lda", predict.type = "response")
lda_model <- mlr::train(lda, myTask)
lda_raw  <- getLearnerModel(lda_model)


lda_scores <- data.frame(
  LD1         = as.numeric(predict(lda_raw, newdata = as.data.frame(getTaskData(myTask)))$x[, 1]),
  Loan_Status = getTaskData(myTask)$Loan_Status
)

ggplot(lda_scores, aes(x = Loan_Status, y = LD1, fill = Loan_Status)) +
  geom_violin(alpha = 0.5) +
  geom_beeswarm(size = 0.8, alpha = 0.6) +
  theme_bw() +
  labs(title = "LDA: LD1 separation by Loan Status")

# 10-fold Repeated CV
cv_desc <- makeResampleDesc("RepCV", folds = 10, reps = 50)

lda_cv   <- resample(lda, myTask, resampling = cv_desc,
                     measures = list(acc, mmce), show.info = FALSE)

calculateConfusionMatrix(lda_cv$pred, relative = TRUE)
caret::confusionMatrix(
  data      = lda_cv$pred$data$response,
  reference = lda_cv$pred$data$truth,
  positive  = "Y"
)

# QDA
qda <- makeLearner("classif.qda", predict.type = "response")

qda_cv <- resample(qda, myTask, resampling = cv_desc,
                   measures = list(acc, mmce), show.info = FALSE)

calculateConfusionMatrix(qda_cv$pred, relative = TRUE)

caret::confusionMatrix(
  data      = qda_cv$pred$data$response,
  reference = qda_cv$pred$data$truth,
  positive  = "Y"
)
```

## Support Vector Machines

```{r Support Vector Machines, message=FALSE, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}

svm <- mlr::makeLearner("classif.svm")

# Defining the hyperparameter space for tuning
svmParamSpace <- makeParamSet(
  makeNumericParam("cost",  lower = 0.1, upper = 10),
  makeNumericParam("gamma", lower = 0.1, upper = 10)
)

# ── Random Search ────────────────────────────────────────────────────────────
randControl  <- makeTuneControlRandom(maxit = 50)
holdout_desc <- makeResampleDesc("Holdout", split = 0.8)

set.seed(42)
svm_rand_tune <- tuneHyperPars(
  learner    = svm,
  task       = myTask,
  resampling = holdout_desc,
  par.set    = svmParamSpace,
  control    = randControl
)

# Hyperparameter impact plot (random search)
rand_effect <- generateHyperParsEffectData(svm_rand_tune)
plotHyperParsEffect(rand_effect,
                    x = "cost", y = "gamma", z = "mmce.test.mean",
                    plot.type = "scatter",
                    colour.limits = c(0, 0.5),
                    palette = "RdBu")

# ── Grid Search ──────────────────────────────────────────────────────────────
gridControl <- makeTuneControlGrid(resolution = 30)

set.seed(42)
svm_grid_tune <- tuneHyperPars(
  learner    = svm,
  task       = myTask,
  resampling = holdout_desc,
  par.set    = svmParamSpace,
  control    = gridControl
)

# Hyperparameter impact plot (grid search)
grid_effect <- generateHyperParsEffectData(svm_grid_tune)
plotHyperParsEffect(grid_effect,
                    x = "cost", y = "gamma", z = "mmce.test.mean",
                    plot.type = "scatter",
                    colour.limits = c(0, 0.5),
                    palette = "RdBu")

# ── Nested CV ────────────────────────────────────────────────────────────────
svm_wrapper <- makeTuneWrapper(
  learner    = svm,
  resampling = holdout_desc,
  par.set    = svmParamSpace,
  control    = gridControl,
  show.info  = FALSE
)

outer_cv <- makeResampleDesc("CV", iters = 5)

set.seed(42)
svm_nested_cv <- resample(svm_wrapper, myTask, resampling = outer_cv,
                           measures = list(acc, mmce), show.info = FALSE)
svm_nested_cv

svm_tuned <- setHyperPars(svm,
                           cost  = svm_grid_tune$x$cost,
                           gamma = svm_grid_tune$x$gamma)

svm_model <- train(svm_tuned, myTask)
```

## Decision Tree

```{r Decision Tree, message=FALSE, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}

tree <- mlr::makeLearner("classif.rpart")

# Defining the hyperparameter space for tuning
treeParamSpace <- makeParamSet(
  makeIntegerParam("minsplit",   lower = 5,    upper = 20),
  makeIntegerParam("minbucket",  lower = 3,    upper = 10),
  makeNumericParam("cp",         lower = 0.01, upper = 0.1),
  makeIntegerParam("maxdepth",   lower = 3,    upper = 10)
)

# ── Hyperparameter tuning (Random Search + 5-fold CV + parallel) ─────────────
cv5_desc     <- makeResampleDesc("CV", iters = 5)
treeControl  <- makeTuneControlRandom(maxit = 200)

parallelStartSocket(cpus = detectCores())

set.seed(42)
tree_tune <- tuneHyperPars(
  learner    = tree,
  task       = myTask,
  resampling = cv5_desc,
  par.set    = treeParamSpace,
  control    = treeControl
)

parallelStop()

# ── Nested CV ────────────────────────────────────────────────────────────────
tree_wrapper <- makeTuneWrapper(
  learner    = tree,
  resampling = cv5_desc,
  par.set    = treeParamSpace,
  control    = treeControl,
  show.info  = FALSE
)

set.seed(42)
tree_nested_cv <- resample(tree_wrapper, myTask,
                            resampling = makeResampleDesc("CV", iters = 5),
                            measures   = list(acc, mmce), show.info = FALSE)
tree_nested_cv

# ── Train final tree ─────────────────────────────────────────────────────────
tree_tuned <- setHyperPars(tree,
                            minsplit  = tree_tune$x$minsplit,
                            minbucket = tree_tune$x$minbucket,
                            cp        = tree_tune$x$cp,
                            maxdepth  = tree_tune$x$maxdepth)

tree_model      <- train(tree_tuned, myTask)
tree_model_data <- getLearnerModel(tree_model)

# Plot the decision tree
rpart.plot(tree_model_data, type = 5, box.palette = "BuBn", fallen.leaves = TRUE)

# Complexity parameter table
printcp(tree_model_data)
```

## Random Forest

```{r Random Forest, message=FALSE, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}

forest <- mlr::makeLearner("classif.randomForest")

# Defining the hyperparameter space for tuning
forestParamSpace <- makeParamSet(
  makeIntegerParam("ntree",    lower = 30, upper = 30),
  makeIntegerParam("mtry",     lower = 6,  upper = 12),
  makeIntegerParam("nodesize", lower = 1,  upper = 5),
  makeIntegerParam("maxnodes", lower = 5,  upper = 20)
)

forestControl <- makeTuneControlRandom(maxit = 200)

# ── Hyperparameter tuning (parallel) ─────────────────────────────────────────
parallelStartSocket(cpus = detectCores())

set.seed(42)
forest_tune <- tuneHyperPars(
  learner    = forest,
  task       = myTask,
  resampling = cv5_desc,
  par.set    = forestParamSpace,
  control    = forestControl
)

parallelStop()

# ── Nested CV ────────────────────────────────────────────────────────────────
forest_wrapper <- makeTuneWrapper(
  learner    = forest,
  resampling = cv5_desc,
  par.set    = forestParamSpace,
  control    = forestControl,
  show.info  = FALSE
)

set.seed(42)
forest_nested_cv <- resample(forest_wrapper, myTask,
                              resampling = makeResampleDesc("CV", iters = 5),
                              measures   = list(acc, mmce), show.info = FALSE)
forest_nested_cv

# ── Train final forest ───────────────────────────────────────────────────────
forest_tuned <- setHyperPars(forest,
                              ntree    = forest_tune$x$ntree,
                              mtry     = forest_tune$x$mtry,
                              nodesize = forest_tune$x$nodesize,
                              maxnodes = forest_tune$x$maxnodes)

forest_model      <- train(forest_tuned, myTask)
forest_model_data <- getLearnerModel(forest_model)

# ── OOB error plot ───────────────────────────────────────────────────────────
oob_error <- forest_model_data$err.rate
n_classes <- ncol(oob_error) - 1  # number of class columns

plot(oob_error[, "OOB"],
     type  = "l",
     ylim  = c(0, max(oob_error, na.rm = TRUE)),
     xlab  = "Number of Trees",
     ylab  = "Error Rate",
     main  = "OOB Error Rate",
     col   = 1)

colors_cls <- 2:(n_classes + 1)
for (i in seq_len(n_classes)) {
  lines(oob_error[, i + 1], col = colors_cls[i])
}

legend("topright",
       legend = colnames(oob_error),
       col    = c(1, colors_cls),
       lty    = 1)

# ── Variable Importance plot ─────────────────────────────────────────────────
randomForest::varImpPlot(forest_model_data)

# ── Information Gain feature filter ─────────────────────────────────────────
filter_data <- generateFilterValuesData(myTask,
                                         method = "FSelectorRcpp_information.gain")
plotFilterValues(filter_data)
```

## Benchmarking

```{r Benchmarking, message=FALSE, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}

# Combine five learners into a list
lda_lrn    <- makeLearner("classif.lda")
qda_lrn    <- makeLearner("classif.qda")

# Re-use the tuned wrappers / tuned learner objects
learner_list <- list(lda_lrn, qda_lrn, svm_tuned, tree_tuned, forest_tuned)

# Benchmark with RepCV (10 folds, 10 reps)
bench_cv <- makeResampleDesc("RepCV", folds = 10, reps = 10)

set.seed(42)
bench_result <- mlr::benchmark(
  learners   = learner_list,
  tasks      = myTask,
  resamplings = bench_cv,
  measures   = list(mmce, acc, fpr, fnr),
  show.info  = FALSE
)

bench_result

# ── Boxplots of accuracy ─────────────────────────────────────────────────────
plotBMRBoxplots(bench_result,
                measure  = acc,
                style    = "box",
                pretty.names = FALSE) +
  aes(colour = learner.id) +
  theme_bw()

# ── Bar chart of average accuracy ────────────────────────────────────────────
bench_df <- as.data.frame(bench_result)

ggplot(bench_df, aes(x = learner.id, y = acc, fill = learner.id)) +
  stat_summary(geom = "bar", fun = "mean") +
  scale_fill_brewer(palette = "YlGnBu") +
  labs(x = "learner.id", y = "acc") +
  theme_bw()
```

## Discussion

**Q1: Why couldn't we use the k-Nearest Neighbors (kNN) algorithm easily on this dataset?**

A1: kNN เป็น instance-based learner ที่ต้องการข้อมูล numeric ทั้งหมด เนื่องจาก dataset นี้มีตัวแปรประเภท factor/categorical หลายตัว (Gender, Married, Education, Self_Employed, Credit_History, Property_Area, Dependents) kNN จึงไม่สามารถคำนวณ distance ได้โดยตรง นอกจากนี้ถ้า encode เป็น integer แล้ว scale ของแต่ละตัวแปรก็ต่างกัน ต้องทำ normalization เพิ่ม และ kNN ยังมีปัญหา curse of dimensionality เมื่อมีหลาย feature

**Q2: In the context of the Dream Housing Finance company, evaluate the importance of the model's Sensitivity versus Specificity. Which metric matters more for their business?**

A2: **Sensitivity (Recall)** สำคัญกว่าในบริบทนี้ Sensitivity วัดว่าโมเดลสามารถจำแนกลูกค้าที่ *ควรได้รับ* สินเชื่อได้ถูกต้องมากน้อยเพียงใด (True Positive Rate) หาก Sensitivity ต่ำ บริษัทจะปฏิเสธลูกค้าที่มีคุณสมบัติครบ ทำให้สูญเสียรายได้จากดอกเบี้ย อย่างไรก็ตาม Specificity (การจำแนกผู้ที่ไม่ควรได้รับสินเชื่อ) ก็มีความสำคัญเพื่อลด NPL (Non-Performing Loan) โดยสรุป บริษัทควร optimize **Sensitivity** เป็นหลัก แต่ต้องระวัง Specificity ไม่ให้ต่ำเกินไปเพื่อควบคุมความเสี่ยงด้านเครดิต

**Q3: Look at the Variable Importance plots generated in Step 9. Which feature was identified as the most dominant predictor of Loan_Status? Does this align with real-world financial logic?**

A3: จากกราฟ Variable Importance และ Information Gain ตัวแปรที่สำคัญที่สุดคือ **`Credit_History`** ซึ่งมีค่า MeanDecreaseGini และ Information Gain สูงกว่าตัวแปรอื่นอย่างชัดเจน สอดคล้องกับความเป็นจริงทางการเงินอย่างมาก เพราะประวัติการชำระหนี้ (credit history) คือตัวชี้วัดหลักที่สถาบันการเงินใช้ประเมินความน่าเชื่อถือของผู้กู้มาโดยตลอด ลูกค้าที่มีประวัติการชำระหนี้ดีย่อมมีความเสี่ยงผิดนัดชำระต่ำกว่า จึงมีโอกาสได้รับอนุมัติสินเชื่อสูงกว่า

------------------------------------------------------------------------

End-of-File\
2026-03-11

```{r}
pwd
```
